Notes from the President

### 2020-09-13 pnsac/newsletter/article_Feature1.tex
First draft for Sep 2020 issue.

\title{Volunteers Assure a Bright Future for Classic Plane}
%\subtitle{Part 4}
\author{Randall Denley}

@@

\end{multicols}

\begin{figure*}[htbp]
	\vspace{2em}
	\centering
	%name of the graphic, without the path AND in EPS format:
	\includegraphics[scale=0.8]{holmgren-timmin-scaled.png}
	%caption of the figure 
	\caption*{\small \em TIm Timmins, left, and Robert Holmgren are forming a
group of skilled volunteers to restore a 55-year-old North Star, the first
Canadian-made plane capable of transcontinental flight.}
	%label of the figure, which has to correspond to \ref{}:
	\label{fig:timmins-holmgren}
\end{figure*}

\begin{multicols}{2}

\textit{This article was published in the Ottawa Citizen on February 1, 2003
under the byline of Randall Denley. It was brought to our attention recently by
Drew Hodge. The article is republished with the express permission of the
Ottawa Citizen, a division of Postmedia Network Inc. (ed.).
}\\

The world's last classic Canadair North Star sits, forlorn, on the tarmac at
the Canada Aviation Museum. It's been parked there since 1966, along with half
a dozen other orphans of the museum, planes that it has no money to restore and
that are too big to bring inside the crowded building.

The paint on the North Star is faded and peeling.Some gaps in the wings are
closed with wire to keep out the birds that have made it home for years.The
engines are protected with pieces of plywood. The interior requires a complete
renovation.

It's a shame to see an important part of Canada's aviation history slide into
this kind of ruin, Thanks to the efforts of two Ottawa men, however, the North
Star finally has a bright future.

Robert Holmgren and Tim Timmins are forming a group of skilled volunteers to
restore the North Star. This kind of volunteer effort will be a first for the
museum, says director general Anthony Smyth, who is supporting the effort,
along with aviation companies including Air Canada and Boeing.

The museum is showing commendable flexibility because the traditional view in
the museum world is that restoration is work that can only be done by trained
experts. Unfortunately, the museum can't afford that kind of work, and that's
one of the reasons the plane has been sitting there for more than 35 years.

This is no small undertaking. The 55-year-old North Star 1-ST is a sizeable
plane, nearly 30 metres long and with a wing span of 35 metres. Restoring it
will be a yearlong process that would cost \$1 million if it were done by paid
staff.

The North Star is an important plane in Canada's aviation history, the first
Canadian-,made plane capable of transcontinental flight. It's considered the
plane that launched the postwar Canadian aviation industry; 70 of them were
built. Brought into service after the Second World War, North Stars carried
passengers for several airlines and troops for the Canadian military.  North
Stars served to airlift Canadian troops to Korea and provided transportation
for VIPs visiting Canada. The plane at the museum belonged to the RCAF.

Holmgren is a retired Air Canada maintenance expert, and Timmins is a former
RCAF navigator who flew on the North Stars. Timmins's former squadron had taken
an interest in the North Star, and Holmgren, who was giving his time as a
volunteer at the museum, pushed it forward.

"We'd like to see it brought back to something like its former self," Holmgren
says.

A temporary structure, perhaps a bubble like the one over the playing field at
Lansdowne Park, will be necessary before the restoration can begin. The plane
is too large to fit inside the museum's existing building and also too big to
be easily transported elsewhere.

Holmgren and Timmins need about 200 volunteers to make the project a success.
Holmgren has approached Air Canada, Bombardier and Boeing to get the help of
past and present employees. About 40 volunteers have already come forward.

The key need is skilled craftsmen with up-to-date large aircraft experience and
who live in or near Ottawa.

Some people who can come in almost every day to run the project are required. 
There is also a demand for people to do record-keeping and research.

The group is hoping to attract national attention for the project, both to
provide volunteer expertise and to help raise money.  This is a plan to help a
national museum, after all.  Smyth says the museum will be involved with a
fundraising campaign, but it's early yet to say how much is required. It
depends partly on how much time and material are donated by corporations. A
full-time salaried project manager will be hired by the museum.

The project will take a conservation approach, retaining as much of the
original material as possible, Smyth says. The restored plane will be
structurally complete and faithful to the original. Returning the plane to
flyable condition is too expensive to be feasible and the museum wouldn't want
to take a chance flying the last remaining version of such an important plane.

Restoring the North Star is just the beginning of what Holmgren and his
associates hope to accomplish. The museum has another half dozen planes outside
that require restoration. Restoring them all will take nine or ten years. The
Museum's new addition, which it hopes to have open by December of this year,
will be large enough to finally bring the North Star indoors.

%If you want to help with the project, contact Robert Holmgren at 748-5972 or
%robertholmgren@rogers.com. The group also has a Web site,
%www.projectnorthstar.ca.

%\address{Web site:
%{\normalfont\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}}\\
%	General enquiries:\email{info@projectnorthstar.ca}}
%

%\begin{quotation}
%	\textit{You have to be careful opening up the panels covering the 
%	engines, you never know if a bird or some animal might have made
%	its home there.}
%\end{quotation}

%\begin{figure*}[htbp]
%	\vspace{2em}
%	\centering
%	%name of the graphic, without the path AND in EPS format:
%	\includegraphics[scale=0.5]{op-hawk-one.jpg}
%	%caption of the figure 
%	%\caption*{\small \em Volunteers put in countless hours restoring a
%	North Star aircraft.}
%	%label of the figure, which has to correspond to \ref{}:
%	\label{fig:op-hawk-one.eps}
%\end{figure*}

>>>>

\title{North Star Propeller Restorations}
%\subtitle{Part 4}
\author{Harry Hope, with assistance from Michael Hope.}

@@

\textit{Hope Aero has played a major role in the restoration of the North Star.
The following article provides some background about the company and its work
on the North Star’s propellers. }\\

Hope Aero was formally government approved to certify aircraft propellers in
1992, at their facility in Mississauga, Ontario. Harry Hope however, already
had 40 years of propeller overhaul experience on many varieties that included
Hamilton Standard propellers, which were quite common, as well as Hartzell,
McCauley, and Sensenich. Harry's first restoration to flying condition, was the
propeller for the Lysander, which was restored by the Air Force in Winnipeg for
their Centennial project in 1967. This Lysander is now displayed at CASM. Since
that time, they have restored several propellers, some to flying condition,
three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
display at the Aviation museum in Boda, Norway.

 Since that time, their capabilities have added– MT, Dowty, Hoffman, and
Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
their customer base expanded to include different types of aircraft, the
requirement for them to service more types became evident. They kept up with
the required technology.

Along the way, they also expanded into aircraft Wheel and Brake overhaul,
sales, service and on wing dynamic balancing of fixed wing and Helicopters.
Harry retired in 2001, but the third generation is now capably looking after
the business.

In the spring of 2006, Hope Aero took on the task of restoring the four North
Star propellers, for the "North Star Restoration" project. The first challenge
was to remove the propellers from the engines as neither the restoration crew
nor Hope Aero had any recent experience removing this type of propeller.
Michael Hope made the trip to Ottawa to assist with this operation and to help
prevent any damage, which turned out to be exceedingly difficult due to the
corrosion, starting with the spinner and then the propeller. Once the propeller
was removed the blades were removed for shipping to Hope Aero.

The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
Aero in 2006. It was completely disassembled, cleaned and inspected. The
external parts were very badly corroded, with most damage being to the aluminum
blades and dome shells, and rust to the steel hub and tail shaft, which was
likely caused by the outside storage environment. The inside parts were in
relatively good condition. It was established, that both technical and visual
effort would be required to restore the propeller to an acceptable display
condition.

The restoration work commenced with the blades, the alcohol de-icer feed shoes,
troughs and paint being removed. The corrosion of the blade surfaces was such
that removal by grinding was abandoned, as the inner half of the blades are
shot peened and grinding would remove this and the authenticity. It was decided
to media blast the complete airfoil surfaces with an aluminum oxide media, thus
removing the surface corrosion but leaving a good surface for paint to be
adhered to; an aviation grade grey paint was applied with yellow stripes at the
tips. New alcohol feed shoes were installed, however the troughs, which direct
the alcohol to the shoes, were not available. As the troughs are hidden inside
the spinner, it was determined that they could be omitted.
The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
tubes, and dome retaining nut, were extremely rusty. They were glass bead
blasted to remove the rust; this did not remove the heavy pitting but did leave
a surface that could be cadmium plated.  After plating, a clear coat material
was sprayed on these parts to improve the resistance to rust. The internal
surfaces were in good condition, and they were well oiled.

A replacement dome shell in relatively good condition was found, requiring only
to be sanded, anodized, and be reinstalled with new seals.

The propeller was then completely assembled to confirm that all the refurbished
parts fit together. The blades were then removed to facilitate being shipping
back to the museum.

All of the four propellers were  found to be in similar condition, and all were
restored in a similar manner.

### 2020-09-13 pnsac/newsletter/article_Feature2.tex
First draft for Sep 2020 issue.

What was undoubtedly true was that Rolls-Royce learned a great deal
about the harsh realities of commercial operation in a short time.
Hives is supposed to have said to TCA, "We didn't know the Merlin until
you started operating it!" In the longer term, this paid off,
TCA---later Air Canada---selecting Rolls--Royce engines for its future
fleet: Darts in Viscounts, Tynes in Vanguards, Conways in DC-8s, and
RB211s in TriStars.

>>>>

What was undoubtedly true was that Rolls-Royce learned a great deal about the
harsh realities of commercial operation in a short time. Hives is supposed to
have said to TCA, "We didn't know the Merlin until you started operating it!"
In the longer term, this paid off, TCA — later Air Canada — selecting
Rolls-Royce engines for its future fleet: Darts in Viscounts, Tynes in
Vanguards, Conways in DC-8s, and RB211s in TriStars.

### 2020-09-13 pnsac/newsletter/article_Feature3.tex
First draft for Sep 2020 issue.

\title{The Whispering Giant}
\author{Dave Lambeth}

@@

\begin{centering}
Twinkle, twinkle, old North Star,\\
With wrinkled skin and creaky
spar,\\
For years you've always been on
sked,\\
As many of our force you've sped\\
to distant lands beyond the skies,\\
you've never failed to see the rise\\
of distant suns or moonrise glare.\\
You've carried us with tender care\\
-- back you came to take us home,\\
from old Japan or France or Rome\\
from England, Goose, and Kef.\\
We cannot hear you.\\
('cause we're deaf)\\
\end{centering}

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{jim_jung_2.eps}
   %caption of the figure 
   \caption*{\small \em A Whispering Giant (North Star 17515) at
     Resolute Bay NWT in about 1963, taken by Radio Operator Jim Jung.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:jim_jung}
\end{figure}

>>>>

\title{North Star Propeller Restorations}
%\subtitle{Part 4}
\author{Harry Hope, with assistance from Michael Hope.}

@@

\textit{Hope Aero has played a major role in the restoration of the North Star.
The following article provides some background about the company and its work
on the North Star's propellers. }\\

Hope Aero was formally government approved to certify aircraft propellers in
1992, at their facility in Mississauga, Ontario. Harry Hope however, already
had 40 years of propeller overhaul experience on many varieties that included
Hamilton Standard propellers, which were quite common, as well as Hartzell,
McCauley, and Sensenich. Harry's first restoration to flying condition, was the
propeller for the Lysander, which was restored by the Air Force in Winnipeg for
their Centennial project in 1967. This Lysander is now displayed at CASM. Since
that time, they have restored several propellers, some to flying condition,
three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
display at the Aviation museum in Boda, Norway.

 Since that time, their capabilities have added– MT, Dowty, Hoffman, and
Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
their customer base expanded to include different types of aircraft, the
requirement for them to service more types became evident. They kept up with
the required technology.

Along the way, they also expanded into aircraft Wheel and Brake overhaul,
sales, service and on wing dynamic balancing of fixed wing and Helicopters.
Harry retired in 2001, but the third generation is now capably looking after
the business.

In the spring of 2006, Hope Aero took on the task of restoring the four North
Star propellers, for the "North Star Restoration" project. The first challenge
was to remove the propellers from the engines as neither the restoration crew
nor Hope Aero had any recent experience removing this type of propeller.
Michael Hope made the trip to Ottawa to assist with this operation and to help
prevent any damage, which turned out to be exceedingly difficult due to the
corrosion, starting with the spinner and then the propeller. Once the propeller
was removed the blades were removed for shipping to Hope Aero.

The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
Aero in 2006. It was completely disassembled, cleaned and inspected. The
external parts were very badly corroded, with most damage being to the aluminum
blades and dome shells, and rust to the steel hub and tail shaft, which was
likely caused by the outside storage environment. The inside parts were in
relatively good condition. It was established, that both technical and visual
effort would be required to restore the propeller to an acceptable display
condition.

The restoration work commenced with the blades, the alcohol de-icer feed shoes,
troughs and paint being removed. The corrosion of the blade surfaces was such
that removal by grinding was abandoned, as the inner half of the blades are
shot peened and grinding would remove this and the authenticity. It was decided
to media blast the complete airfoil surfaces with an aluminum oxide media, thus
removing the surface corrosion but leaving a good surface for paint to be
adhered to; an aviation grade grey paint was applied with yellow stripes at the
tips. New alcohol feed shoes were installed, however the troughs, which direct
the alcohol to the shoes, were not available. As the troughs are hidden inside
the spinner, it was determined that they could be omitted.

The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
tubes, and dome retaining nut, were extremely rusty. They were glass bead
blasted to remove the rust; this did not remove the heavy pitting but did leave
a surface that could be cadmium plated.  After plating, a clear coat material
was sprayed on these parts to improve the resistance to rust. The internal
surfaces were in good condition, and they were well oiled.

A replacement dome shell in relatively good condition was found, requiring only
to be sanded, anodized, and be reinstalled with new seals.

The propeller was then completely assembled to confirm that all the refurbished
parts fit together. The blades were then removed to facilitate being shipping
back to the museum.

All of the four propellers were found to be in similar condition, and all were
restored in a similar manner.

### 2020-09-13 pnsac/newsletter/article_MemberInterview.tex
First draft for Sep 2020 issue.

\title{Our Members}
\author{Interview with Garry Dupont}

@@

\end{multicols}

@@

\begin{multicols}{2}

1.\textit{What is your background in aviation?}

I graduated from the Confederation College Aircraft Maintenance course
in the Spring of 1975. After returning to my home town in the Red Lake
District I gained employment with Red Lake Seaplane Service. After
completing the course I had to apprentice for 18 months before writing
the final exam for my Aircraft Maintenance Engineers' License. I
successfully wrote my final exam in December 1976 and obtained my AME
certification. In February 1977 I was hired by Perimeter Aviation in
Winnipeg Manitoba. There I obtained my Turbine Engine and
Pressurization Aircraft endorsements.  I started my 30 year career with
the RCMP Air Services Branch in February 1981. My career took me to
Edmonton Alberta, Goose Bay Labrador, Winnipeg Manitoba then finally to
Ottawa. I finished my career as Quality Assurance Manager.

2. \textit{How long have you been involved with Project North Star and how and
why did you get involved?}

I attended the very first meeting held for Project North Star. This was
at my wife's suggestion that it would be a good retirement pastime.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
%   %caption of the figure 
%   \caption*{\small \em Forward cockpit -- complete.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:fwd_cockpit_complete}
%\end{figure}

3. \textit{What has been the history of your involvement to date?}

While I was employed I worked an evening shift every 4 weeks so I would
come in to CASM in the morning for a few hours. Upon my retirement Mike
Irving had me work in the engine shop, which I have been doing since my
retirement. I also have been on the board of directors for a number of
years.

%\begin{figure}[htbp]
%	\vspace{2em}
%	\centering
%	%name of the graphic, without the path AND in EPS format:
%	\includegraphics[scale=0.9]{resources/PNS2008-10-26_23-31-42.png}
%	%caption of the figure 
%	\caption*{\small \em Propeller assembly.}
%	%label of the figure, which has to correspond to \ref{}:
%	\label{fig:propeller}
%\end{figure}

%\begin{figure}[htbp]
%	\vspace{2em}
%	\centering
%	%name of the graphic, without the path AND in EPS format:
%	\includegraphics[scale=1.3]{resources/NorthStarCockpitDec2011.png}
%	%caption of the figure 
%	\caption*{\small \em Forward cockpit -- complete.}
%	%label of the figure, which has to correspond to \ref{}:
%	\label{fig:fwd_cockpit_complete}
%\end{figure}

4. \textit{What has been the highlight of your involvement?}

The biggest highlight for me was installing number 2 engine. It was
about half completed when I started in the engine shop. This is the
largest engine I have ever installed in my entire career.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.75]{resources/NavigatorRack.png}
%   %caption of the figure 
%   \caption*{\small \em Navigator rack.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:nav-radio-rack}
%\end{figure}

5. \textit{What has been the most challenging part of your involvement?}

One of the most challenging parts was learning how the engine worked
with all its systems. I found a RCAF training manual in the office and
took it home. Over several months of reading I started to understand
what made this monster of old technology tick. Another challenge was
learning the British Hardware system and tools. A further challenge was
reading  the old manuals which were written before they were
standardized by the ATA chapter system which all aircraft manuals
follow today. Learning the art of conservation was also a  challenge
but was made easier by the guidance of Mike Irving.

>>>>

\title{Our Members}
% \author{Interview with Garry Dupont}

@@

% \end{multicols}

@@

% \begin{multicols}{2}

Readers of the Chronicle will know that this item usually refers to one of our
members. This edition, however, salutes all of our members who worked as
volunteers on the restoration of the North Star engines namely:

\begin{itemize}
  \item Charles Baril 
  \item John Corby (deceased)
  \item Ted Devey (deceased)
  \item Ed Hogan
  \item Peter Houston
  \item Ron Lemieux
  \item Richard Lodge
  \item Wib Neal
  \item Jim Riddoch
  \item Stan Rideout
  \item Ted Slack
  \item John Tasseron
  \item Bill Tate
  \item Tim Timmins
  \item Giuseppe Zanetti
\end{itemize}

The following members have provided comments on their personal experiences.

\textbf{Charles Baril}

Working on all 4 Merlin engines has been an extremely rewarding experience that
has taught me a lot about older technology, an appreciation for history
involved (and those in the engine crew with that knowledge) and learning new
skills and techniques. Although most of the parts were auxiliary systems,
seeing them as part of the full engine gives a sense of accomplishment. It is
very satisfying to see a part going from the dirty, oil covered, rusty part to
newly refurbished one that looked like it was overhauled in an RCAF maintenance
depot. It has been an honour to be part of this project and restoration of
these pieces of history.

In particular I remember working on the lower cowling flaps.  Trying to undo
the inner skin was difficult due to longer rivets through the mounts for the
Flaps. It took several tries and a few extra holes, but it was done. Fixing
those holes often meant using original methods which was revealing and a
commitment to historical methods.

I also recall the work on the filter ducts.  The filter duct is the last stage
of the air intake before the supercharger. During its cleaning, there was a
starling nest on top of one filter. This caused a lot of corrosion damage that
needed to be removed and treated. Yet, what got me was that starlings could get
that far into the engine.

\textbf{Richard Lodge}

As probably the only volunteer who ever worked in the Aero Engine Division of
Rolls-Royce in Derby, England, I was greatly interested in becoming involved in
Project North Star, even though, as an accountant, I had never worked on the
design or building of an aeroengine.  By the time I started work at Rolls-Royce
in 1963, new build of Merlin engines had ceased, and only spare parts were
being manufactured.

I have several strong memories of working on the Merlin engine.  In general,  I
particularly remember John Tasseron working endlessly on freeing seized piston
rings.  I learned much from John about patiently working at something which
appeared to be impossible! Later, when working with  R\'{e}j Demers and Garry
Dupont, I started to learn how to rivet and how to lock wire. The most
important lesson for me has been, photograph, photograph, photograph, at each
stage of dismantling, cleaning and reinstalling an engine part. When sitting in
an overcrowded office at Rolls-Royce, it never occurred to me that at some time
in the future I would actually be working on a Merlin engine or working my way
through an original parts manual to work out how to reassemble an item.  Life
is certainly full of interesting surprises.

\textbf{John Tasseron}
 
The list below is intended to give a brief overview of the various highlights
that arose during my participation in the restoration of the second, third and
fourth Merlin engines:

\begin{enumerate}
  \item It's not every day one gets the opportunity to see up close what makes
the Rolls-Royce 620 Series Merlin engine tick!
  \item It's a real challenge to safely lower each "power egg" down off the
wing, place it in "the big blue stand" and move it in to the workshop.
  \item Then starts the process of carefully separating about 11,000 individual
parts before cleaning, inspecting and re-conditioning them as necessary. Do not
lose any and make sure everything is photographed for museum records!
  \item When things unexpectedly go wrong, the real expertise kicks in to
perhaps develop a repair scheme or even manufacture a brand-new replacement
part. The objective is to bring everything back together again and end up with
a "like new" engine, ready for preservation.
  \item It all takes team effort based on effective communication and
co-operation. There are no dumb questions and the answers are all aimed at the
challenge to create an "airworthy" product that will never fly.
  \item Finally, its important and fun to share the project with a curious
public, so be ready to answer any questions that might be asked about the
aircraft and its engines!
\end{enumerate}

%1.\textit{What is your background in aviation?}
%
%
%2. \textit{How long have you been involved with Project North Star and how and
%why did you get involved?}
%
%3. \textit{What has been the history of your involvement to date?}
%
%4. \textit{What has been the highlight of your involvement?}
%
%5. \textit{What has been the most challenging part of your involvement?}

### 2020-09-13 pnsac/newsletter/article_PresReport.tex
First draft for Sep 2020 issue.

Since the last issue of the NStar Chronicle we have taken part in the
70th Anniversary celebrations of the first non-stop transcontinental
flight across Canada which was flown by a North Star. We partnered with
the Museum for this event and were very pleased to welcome
representatives from the RCAF who joined us for the day. After the
formal part of the celebration, when impressive plaques of the aircraft
made by our member Chris McGuffin, had been presented to the RCAF and
the Museum Director General, Chris Kitzan, everybody moved to the
Reserve Hangar and the aircraft was opened to our association members,
guests and in the afternoon to 90 members of the public who had
pre-booked a tour. R\'{e}jean(R\'{e}j)Demers, the Museum Special Project
Manager, guided the tours assisted by several of our members. Towards
the end of the tour session someone in the main Museum activated the
fire alarm. The Museum was promptly evacuated, and the public was
recommended to go to the Reserve  Hangar until the fire issue had been
dealt with (it turned out to be a false alarm). Quite unexpectedly R\'{e}j
and the guides found themselves surrounded by many Museum visitors who
had not signed up for the North Star tour. Unperturbed by this event,
Réj and the guides welcomed the unscheduled visitors and gave them an
impromptu tour of the aircraft. Both our treasurer and our membership
secretary were pleased with the days' result!

As I write this the days are getting longer and R\'{e}j is planning work to
be done on the aircraft when it is moved outside onto the ramp in the
spring. Each year we do work on the aircraft which cannot be done
inside the Reserve Hangar during the winter. We also prepare the
aircraft for display when, once again, we open the aircraft to enable
the public to view progress made in the last year.

This year we are planning two major display events. The first will be
on the Doors Open Ottawa weekend (June 6 and 7). This is a great time
for our members to meet the public, many of whom have an interest in
aviation. Our second main event, on Canada Day, is enjoyable for all
the members who sign up to be part of the team meeting the public.
There are always large crowds, many of them family groups, If any of
you reading this article would be interested in joining our members for
either of these events please contact me (my co-ordinates are on the
back page of this issue of the Chronicle)

Shortly before Christmas a major milestone, in the  restoration of the
North Star, was reached. Work on Merlin engine \#4 was completed and it
was dragged on its cradle by the Museum forklift truck to the
Restoration hangar for later mounting on the aircraft to join the other
three restored engines. Chris McGuffin put some video clips onto our
Facebook page of the engine being dragged through the Museum parking
lot and into the hangar. Watch for more of our Facebook messages since
Chris frequently posts new items.

In the late fall we welcomed the appointment of Erin Secord as the
Museum's new Manager Conservation Services. She is responsible also for
conservation of the collections at the Science and Tech Museum. Erin
has now been able to join us at two of our Members' meetings.

Finally, we are now working to update the Memorandum of Understanding
between PNSAC and the Museum.  Much has changed since the original
agreement was signed in May 2006.

Our next Members' Meeting is scheduled for Saturday, June 20, 2020
which will also be our AGM. Do reserve this date in your calendar if
you can be in the Ottawa area on that day. The meeting will be held at
the Canada Aviation and Space Museum. More details will follow later.

>>>>

As always, before I start to write the Notes from the President, I reread my
previous Notes to remind myself of what I had previously said in order to set
the record straight if necessary. As I write this at the beginning of
September, I am now aware of the impact of the COVID-19 pandemic on the
restoration schedule for the North Star and the operations of your
Association.

Since March, restoration work on the aircraft has been suspended and the
activities of the Association have been limited. We have, however, been keeping
in touch with each other through the weekly Zoom meeting organized by the
Museum's volunteer co-coordinator Cédric St.-Amour. This virtual meeting takes
place every Thursday at 10 AM EST; it is open to all our volunteers and any
other PNSAC members who would like to join in.

As previously reported, the North Star project reached a major milestone with
the completion of the restoration of Engine \#4 which was moved into the
reserve hangar on a cold December 2019 morning.

Since we are unable to physically meet and there has been no restoration work
carried out, we felt that a special issue of the NSTAR Chronicle devoted
entirely to the 15 year restoration of all four of the Merlin engines would be
of interest to members.

Understandably, with a project which has taken many years. the volunteers
working on the engines have changed over time. Some have retired or stopped
volunteering and sadly, some have died. Earlier this year we asked Ted Devey to
contribute a piece but before writing about of his work, he passed away. Ted
was involved with restoring all the engines and for several of the early years
was the volunteer leading the restoration team. Ted was a great raconteur and
our oldest volunteer at 93 years old. His recollections would have added
considerably to the interest of this special issue of the NSTAR
Chronicle.

After much consideration, the Museum has decided that volunteering cannot
restart before January 2021. Until then we will continue with our regular Zoom
meetings and will also organize a virtual Annual General Meeting for the
Association. We will also continue to provide updates on the situation and much
will depend upon how the different levels of government handle the ongoing
recovery from the pandemic.

There is an old saying "That it is an ill wind that blows nobody any good."
This is very true even of this pandemic. One of the very pleasant results is
having conversations with PNSAC members who don' t reside in the National
Capital area and have sent membership renewals by mail. Where a renewal has
been received and confirmation has been delayed as a result of the pandemic, I
try to call a member on the telephone to explain what is happened and that the
renewal has not been lost. As a result I have had many interesting phone calls
with people whom I have never spoken to before. A definite bonus for me.
